A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-10-09 paper/paper.tex
finished RL-semantics

\begin{abstract}
The default mode network (DMN) is the neural representation of the human
baseline mental activity.
%
Many research streams agree on its likely role in evolutionarily adaptive
envisioning of past and hypothetical scenarios to predict the environment.
The DMN would hence be dedicated to continuous autobiographical memory retrieval,
generation of hypothetical outcomes,
and reward contingency evaluation when letting the mind go.
It explains its highest energy consumption in the brain and
its intimate coupling with conscious awareness.
%
This concept paper proposes a process model that describes
\textit{how} the DMN may actually implement continuous
environmental assessment and prediction.
DMN function is recast in mathematical terms
from the perspective of control theory and
reinforcement learning.
Neurobiological evidence on the human and animal DMN
is linked to recent progress in autonomous systems research
in the machine-learning domain.
We argue that this engineering approach to DMN function can parsimoneously
explain a variety of existing neuroscientific hypotheses.
%
Formal models of the computational processes subserved in the DMN
could offer statistical control of predictive human behavior.

@@

The HC is well known to be involved in memory and
spatial navigation in animals and humans
(von Bechterew, 1900; Scoville and Milner, 1957; O’Keefe and Dostrovsky, 1971; Maguire et al., 2000).
Its highly recursive anatomical architecture
may be specifically designed to allow reconstructing
entire episodes of experience from memory fragments.
%
While the HC in the medial temporal lobe system
is traditionally believed to allow remembering the past,
there is now increasing evidence for a role
in constructing mental models in general (Schachter/Buckner;
Maguire 2016).
Indeed,
hippocampal damage is
not only associated with an impairment in reexperiencing the past (i.e., amnesia),
but also thinking about one’s own future and
imagining new fictitious experiences more broadly (Hassabis/PNAS).
Mental scenes created by hippocampus-lesioned patients exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal hippocampus revealed
some constantly active neuronal ensembles whose firing coincided with
distinct locations in space while the animal navigated through its surroundings.
London taxi drivers, individuals with high performance in spatial navigation,
were shown to exhibit increased grey matter volume in the
posterior hippocampus (Maguire et al. 2006b).
But encoding and generative reconstruction in the hippocampus extends
beyond mere spatial knowledge of the environment.
Based on large-scale recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events (buszaki2004 theta oscillations).
Specific spiking sequences, that were elicited in experimental task conditions,
have been shown to be reenacted spontaneously during sleep.
Moreover, spike sequences measured in hippocampal place cells of rats
featured reoccurrence directly after experimental trials
as well as directly before upcoming experimental trials (Diba + Buzsaki 2007).
Such hippocampal ensemble burts during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the DMN.
These mechanisms of the hippocampus probably make important contributions to the
recollection of autobiographical memory episodes and other
reexperienced or newly generated mental scenarios (Hassabias2007).
%
The HC thus orchestrates elements of experienced environmental aspects for
consolidations based on reenactment and for integration into 
rich mental scene construction. In this view, the HC my even influence
ongoing perception in the current environment (Maguire 2016).
-> self-play in deepmind Go
-> Kai had an interesting idea: the deja-vu effect could be when you already accidentally perturbed past such that it really occures in reality later

@@

The left TPJ in turn exhibits a close topographical relationship to
Wernicke's area
involved in the comprehension or understanding of written and spoken language.
Neurological patients with damage caused to Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Their speech
preserves natural rhythm and about normal syntax, yet the
voiced sentences are devoid of meaning (i.e., aphasia).
Abstracting from the typical semantic interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integrating of world knowlege.
For instance, LTPJ lesions also entail problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing any language content
(Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Also inner speech hinges on knowledge retrieval of statistical structure
about the physical and inter-personal world.
Multivariate volumetric analyses showed that the internal production of
formulated thought ("language of the mind") is closely related to the left TPJ
(Geva et al., 2011).
A number of neuroimaging studies demonstrated LTPJ activity responses
to sementically aberrant words in sentences, which suggests a role
in binding a given semantic concept into an integrated whole.
Further,
episodic memory recall and imagination strongly draw on
complex world knowledge.
Isolated building blocks of world statistics probably get reassembled
in internally generated visual scenarios that
navigate present action, weigh hypothetical possibilities, and forcast the future.
%
The LTPJ may hence facilitate the automated prediction of events
by incorporating experience-derived models of the world
into ongoing action, planning, and problem solving.
world -> policy matrix driven computation
self -> value matrix driven computation

@@

\paragraph{TODO: Monte Carlo optimization.}
Some text snippets that might be pertinent:
Monte-Carlo simulation can be used to update the agent’s belief state
\cite{silver2010monte}.
Monte-Carlo simulation provides a simple method for evaluating
a state s.
In any MDP, there is at least one optimal policy matrix that
achieves the optimal value function???
The simulator is used to generate sequences of states, observations and rewards. These simulations are used to update the value function, without ever looking inside the black box describing the model’s dynamics. In addition, Monte-Carlo methods have a sample complexity that is determined only by the underlying difficulty of the POMDP, rather than the size of the state space or observation space.
Monte-Carlo simulation can be turned into a simple control algorithm by evaluating all legal actions and selecting the action with highest evaluation [15]. Monte-Carlo simulation can be extended to partially observable MDPs.
In these challenging POMDPs, Monte-Carlo simulation provides an effective mechanism both for tree search and for belief state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible.
Link this to experience sampling in the hippocampus.
Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines.

@@

\subsection{Semantic Hypothesis}
Another frequently proposed account of DMN function revolves
around forming meaning associations and analogies between
the currently experience and
abstract knowledge derived from past experiences
\cite{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge extractable from past memories
\cite{bar2009proactive}.
%
These semantic analogies can be used to mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future analogical associations.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \cite{addis2008age, spreng2006temporal}.
A constructive account further explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
It is also closely related to episodic memory, language, problem solving,
planning, estimating others' thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\cite{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \cite{tolman1948cognitive}.

\begin{itemize}
  \item The notion of semantic or knowledge association is
  incorporated into the MDP as the Markov property,
  that is, the current state directly results from the
  agent's history of states and actions. The learned
  value matrix and action transition schedules drive
  stimulus processing and action choice in the present.
  \item In both, semantic hypothesis and MDPs,
  there is no evidence to indicate that predictions of various
  levels of complexity, abstraction, timescale and purpose
  use mechanisms that are qualitatively different. This concurs with
  DMN activity increases across time, space, and content domains
  demonstrated in many neuroimaging studies. The semantic hypothesis
  and RL account provide explanations why hippocampus lesion does
  not only impair retrieving memories, but also hypothetical and future
  thinking \cite{hassabis2007patients}.
  \item Both the semantic hypothesis and MDP expose mechanisms of
  how alternative decision trees could be mentally explored.
  \item While mental scene construction implies a distinction between
  environmental perception and internally generated mind-wandering
  \footnote{``A computational model of how such a process might be structured
  is far from being defined, but it will probably require a form of
  regulation by which perception of the current world is suppressed
  while simulation of possible alternatives are constructed,
  followed by a return to perception of the present.''\cite{buckner2007self}},
  the MDP naturally integrates the former egocentric
  (related to current action, state and reward) and the later
  allocentric (related to past and future actions, states, and rewards)
  angles on the world in a same optimization problem.
  \item The semantic account of DMN function does not offer 
  a mechanistic explanation how semantic world knowledge and analogies thereof
  lead to prediction of future actions and states.
  \item The semantic hypothesis does not explain why memory recall
  for scene construction in humans is fragmentary and noisy
  instead of accurate and reliable.
  Yet, the MDP framework provides an algorithmic explanation in that
  stochasticity of the parameter space search implemented
  by Monte Carlo solvers provably yields better models of the world.
  That is, the purposeful stochasticity of policy and value estimation
  in MDPs provides a candidate explanation for why humans
  have evolved imperfect memories as the better adaptation.
  \item While both semantic and MDP account propose memory-based internally
  generated information for probabilistic mental models of action outcomes,
  only MDPs render explicit the grounds on which the final action is
  actually chosen (i.e., the estimated cumulative reward).
  \item In sum, episodic scene construction according to the semantic
  hypothesis lacks an explicit time and incentive model.
\end{itemize}

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constant, high energy costs?
MDPs provide an attractive
computational model how the human association cortex
might implement high-level control of the environment to
optimize the organism's fate by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN literature by a simple but non-trivial concept.
%
Viewed as a MDP, behavior unfolds by integrating across past
and future experience to guide action choice in the present context.
This account is more compatible with the DMN's involvement across
autobiographical memory retrieval,
abstract reasoning based on internally generated scenes,
and self-projection into the expected future.
MDPs provide a mathematical formalism how
optimal substructure can be recursively exploited
when contemplating nested decisions.
Algorithmic gains by injecting stochasticity in past
behavior and outcomes may plausibilize why
very accurate memories have been evolutionarily de-selected
and why human creativity might be adaptive.
%
In principle,
neuroscientific experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of RL systems. This makes the proposed machine-learning
approach to DMN function not only practically computable but
also neuroscientifically falsifiable.
At the least, we propose an alternative vocabulary to
describe and discuss neuroscientific studies in DMN research.
%
Ultimately,
the DMN can be viewed as a smoothing kernel extending
from the relatively recent past to the relatively close future
to constantly compute complex gain opportunities in the world.

>>>>

\begin{abstract}
The default mode network (DMN) is the neural representation of the human
baseline mental activity.
%
Many research streams agree on its likely role in evolutionarily adaptive
envisioning of past and hypothetical scenarios to predict the environment.
The DMN would hence be dedicated to continuous autobiographical memory retrieval,
generation of hypothetical outcomes,
and reward contingency evaluation when letting the mind go.
It explains its highest energy consumption in the brain and
its intimate coupling with conscious awareness.
%
This concept paper proposes a process model that describes
\textit{how} the DMN may actually implement continuous
environmental assessment and prediction.
DMN function is recast in mathematical terms
from the perspective of control theory and
reinforcement learning.
Neurobiological evidence on the human and animal DMN
is linked to recent progress in autonomous systems research
in the machine-learning domain.
We argue that this engineering approach to DMN function can parsimoneously
explain a variety of existing neuroscientific hypotheses.
%
Formal models of the computational processes subserved in the DMN
could offer access routes to statistical optimality
of human behavior.

@@

The HC is well known to be involved in memory and
spatial navigation in animals and humans
(von Bechterew, 1900; Scoville and Milner, 1957; O’Keefe and Dostrovsky, 1971; Maguire et al., 2000).
Its highly recursive anatomical architecture
may be specifically designed to allow reconstructing
entire episodes of experience from memory fragments.
%
While the HC in the medial temporal lobe system
is traditionally believed to allow remembering the past,
there is now increasing evidence for a role
in constructing mental models in general (Schachter/Buckner;
Maguire 2016).
Indeed,
hippocampal damage is
not only associated with an impairment in reexperiencing the past (i.e., amnesia),
but also thinking about one’s own future and
imagining new fictitious experiences more broadly (Hassabis/PNAS).
Mental scenes created by hippocampus-lesioned patients exposed a lack of
spatial integrity, richness in detail, and overall coherence.
%
Single-cell recordings in the animal hippocampus revealed
some constantly active neuronal ensembles whose firing coincided with
distinct locations in space while the animal navigated through its surroundings.
London taxi drivers, individuals with high performance in spatial navigation,
were shown to exhibit increased grey matter volume in the
posterior hippocampus (Maguire et al. 2006b).
But encoding and generative reconstruction in the hippocampus extends
beyond mere spatial knowledge of the environment.
Based on large-scale recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events (buszaki2004 theta oscillations).
Specific spiking sequences, that were elicited in experimental task conditions,
have been shown to be reenacted spontaneously during
quiet wakefulness and sleep \cite{hartley2014space}.
Moreover, spike sequences measured in hippocampal place cells of rats
featured reoccurrence directly after experimental trials
as well as directly before upcoming experimental trials (Diba + Buzsaki 2007).
Such hippocampal ensemble burts during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the DMN.
These mechanisms of the hippocampus probably make important contributions to the
recollection of autobiographical memory episodes and other
reexperienced or newly generated mental scenarios (Hassabias2007).
%
The HC thus orchestrates elements of experienced environmental aspects for
consolidations based on reenactment and for integration into 
rich mental scene construction. In this view, the HC my even influence
ongoing perception in the current environment (Maguire 2016).
-> self-play in deepmind Go
-> Kai had an interesting idea: the deja-vu effect could be when you already accidentally perturbed past such that it really occures in reality later

@@

The left TPJ in turn exhibits a close topographical relationship to
Wernicke's area
involved in the comprehension or understanding of written and spoken language.
Neurological patients with damage caused to Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Their speech
preserves natural rhythm and about normal syntax, yet the
voiced sentences are devoid of meaning (i.e., aphasia).
Abstracting from the typical semantic interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integrating of world knowlege
to model action concepts
\cite{binder2011neurobiology, seghier2013angular}.
For instance, LTPJ lesions also entail problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing any language content
(Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Also inner speech hinges on knowledge retrieval of statistical structure
about the physical and inter-personal world.
Multivariate volumetric analyses showed that the internal production of
formulated thought ("language of the mind") is closely related to the left TPJ
(Geva et al., 2011).
A number of neuroimaging studies demonstrated LTPJ activity responses
to sementically aberrant words in sentences, which suggests a role
in binding a given semantic concept into an integrated whole.
Further,
episodic memory recall and imagination strongly draw on
complex world knowledge.
Isolated building blocks of world statistics probably get reassembled
in internally generated visual scenarios that
navigate present action, weigh hypothetical possibilities, and forcast the future.
%
The LTPJ may hence facilitate the automated prediction of events
by incorporating experience-derived models of the world
into ongoing action, planning, and problem solving.
world -> policy matrix driven computation
self -> value matrix driven computation

@@

\paragraph{TODO: Monte Carlo optimization.}
Some text snippets that might be pertinent:
Monte-Carlo simulation can be used to update the agent’s belief state
\cite{silver2010monte}.
Monte-Carlo simulation provides a simple method for evaluating
a state s.
In any MDP, there is at least one optimal policy matrix that
achieves the optimal value function???
The simulator is used to generate sequences of states, observations and rewards. These simulations are used to update the value function, without ever looking inside the black box describing the model’s dynamics. In addition, Monte-Carlo methods have a sample complexity that is determined only by the underlying difficulty of the POMDP, rather than the size of the state space or observation space.
Monte-Carlo simulation can be turned into a simple control algorithm by evaluating all legal actions and selecting the action with highest evaluation [15]. Monte-Carlo simulation can be extended to partially observable MDPs.
In these challenging POMDPs, Monte-Carlo simulation provides an effective mechanism both for tree search and for belief state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible.
Link this to experience sampling in the hippocampus.
Much recent work points to Monte Carlo or stochastic sampling– based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. 
For sampling-based methods it is still almost non-existent beyond the description of receptive fields.

@@

\subsection{Semantic Hypothesis}
Another frequently proposed account of DMN function revolves
around forming meaning associations and analogies between
the currently experience and
abstract knowledge derived from past experiences
\cite{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge extractable from past memories
\cite{bar2009proactive}.
%
Going much beyond language ifself,
semantic analogies can be used to mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future analogical associations.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \cite{addis2008age, spreng2006temporal}.
A constructive account further explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
It is also closely related to episodic memory, language, problem solving,
planning, estimating others' thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\cite{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \cite{tolman1948cognitive}.

\begin{itemize}
  \item The notion of semantic or knowledge association is
  incorporated into the MDP as the Markov property,
  that is, the current state directly results from the
  agent's history of states and actions. The learned
  value matrix and action transition schedules drive
  stimulus processing and action choice in the present.
  \item In both, semantic hypothesis and MDPs,
  there is no evidence to indicate that predictions of various
  levels of complexity, abstraction, timescale and purpose
  use mechanisms that are qualitatively different. This concurs with
  DMN activity increases across time, space, and content domains
  demonstrated in many neuroimaging studies. The semantic hypothesis
  and RL account provide explanations why hippocampus lesion does
  not only impair retrieving memories, but also hypothetical and future
  thinking \cite{hassabis2007patients}.
  \item Both the semantic hypothesis and MDP expose mechanisms of
  how alternative decision trees could be mentally explored.
  \item While mental scene construction implies a distinction between
  environmental perception and internally generated mind-wandering
  \footnote{``A computational model of how such a process might be structured
  is far from being defined, but it will probably require a form of
  regulation by which perception of the current world is suppressed
  while simulation of possible alternatives are constructed,
  followed by a return to perception of the present.''\cite{buckner2007self}},
  the MDP naturally integrates the former egocentric
  (related to current action, state and reward) and the later
  allocentric (related to past and future actions, states, and rewards)
  angles on the world in a same optimization problem.
  \item The semantic account of DMN function does not offer 
  a mechanistic explanation how semantic world knowledge and analogies thereof
  lead to prediction of future actions and states.
  \item In contrast to existing accounts on semantics and
  mental scene construction, the random and creative aspects of DMN function
  are here directly related to the advantages of stochastic optimization.
  \item The semantic hypothesis does not explain why memory recall
  for scene construction in humans is fragmentary and noisy
  instead of accurate and reliable.
  Yet, the MDP framework provides an algorithmic explanation in that
  stochasticity of the parameter space search implemented
  by Monte Carlo solvers provably yields better models of the world.
  That is, the purposeful stochasticity of policy and value estimation
  in MDPs provides a candidate explanation for why humans
  have evolved imperfect memories as the better adaptation.
  \item While both semantic and MDP account propose memory-based internally
  generated information for probabilistic mental models of action outcomes,
  only MDPs render explicit the grounds on which the final action is
  actually chosen (i.e., the estimated cumulative reward).
  \item In sum, episodic scene construction according to the semantic
  hypothesis lacks an explicit time and incentive model.
\end{itemize}

@@

\section{Conclusion}
What brain function could be important enough
for the existence and survival of the human species
to warrant constant, high energy costs?
MDPs provide an attractive
computational model how the human association cortex
might implement supra-modal control of the environment to
optimize the organism's fate by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN literature by a simple but non-trivial concept.
%
Viewed as a MDP, behavior unfolds by integrating across past
and future experience to guide action choice in the present context.
This account is more compatible with the DMN's involvement across
autobiographical memory retrieval, problem solving,
abstract reasoning based on internally generated scenes, social cognition,
as well as delay discounting and self-related prospection.
MDPs provide a mathematical formalism how
optimal substructure can be recursively exploited
when confronted with nested decisions.
Algorithmic gains by injecting stochasticity in past
behavior and outcomes may plausibilize why
very accurate memories have been evolutionarily de-selected
and why human creativity might be adaptive.
%
In principle,
neuroscientific experiments can be designed that operationalize
the set of action, value, and state variables that determine
the behavior of intelligent RL systems.
The proposed machine-learning
approach to DMN function is not only practically computable but
also neuroscientifically falsifiable.
At the least, we propose an alternative vocabulary to
describe and discuss neuroscientific studies in DMN research.
%
Ultimately,
the DMN can be viewed as a smoothing kernel extending
from the relatively recent events to potentially upcoming events
to constantly improve sensory input and peformed action in the world.
